\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-09-23 demystify.tex

Then, applications are reviewed by a \textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note GMUU, UMass, and probably many other schools, have a separate committee for the MS program.

@@

In many cases, the admission committee involves assistant professors in the department. This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

\begin{commentbox}[Vu] 
We usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own (e.g.,
supported by your government or having external grants). We also justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs). 
\end{commentbox}

@@

Several things about stipends.  First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like most source of income in the US, you will have to pay tax on your stipend.  Fourth, CS department typically has higher stipend comparing to other study fields.  Finally, private universities might pay more for stipend (but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour).

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount set by the college, but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof. 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

A related topic is to consider is how to refer to a professor you know. Here, because you are already familiar with these individuals, you can informally use their firstname (or Dr./Prof., if you want to be formal) as mentioned above. You can also include their institution if it makes it more precise.  For example, I did my postdoc with Jeff Foster at Univ. of Maryland.   Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many students, especially Vietnamese, include a lengthy title of people they know, e.g., my advisors are Asst. Prof. X and Assoc. Prof. Y., and I also collaborate with ACM Distinguished Scientist Z.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. In short, these nuances represent some cultural differences that you may encounter and will gradually adapt to (more on cultural differences in \S\ref{sec:cultural}).

>>>>

Then, applications are reviewed by a \textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note that GMU, UMass, and probably many other schools, have separate committees for MS programs.

@@

In many cases, the PhD admission committee involves assistant professors in the department. This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own (e.g.,
supported by your government or having external grants). We also justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs).

@@

Several things about stipends.  First, the amount of stipend depends on the university, which in turn depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is typically 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like most source of income in the US, you will have to pay tax on your stipend.  Fourth, CS department typically has higher stipend comparing to other study fields.  Finally, private universities might pay more for stipend (but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour).

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount set by the college, but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof. 
\tcblower
Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\paragraph{Referring to professors you know} Here, because you are already familiar with these individuals, you can just informally use their firstname (or Dr./Prof., if you want to be formal) as mentioned above. You can also include their institution if it makes it more precise.  For example, I did my postdoc with Jeff Foster at Univ. of Maryland.   Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., my advisor is Asst. Prof. X, and I also collaborate with ACM Distinguished Scientist Y.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. In short, these nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.
